
---
title: 'Logistic Regression'
author: "Adam Hendel"
date: "03/13/2018"
output: word_document
fontsize: 12pt
---

Create a Word document from this R Markdown file for the following exercises.  Submit the R markdown file and resulting Word document via D2L Dropbox.   

## Exercise 1

A study was conducted whereby the type of anesthetic (A or B), nausea after the surgery (Yes or No), the amount of pain medication taken during the recovery period, and age for a random sample of 72 patients undergoing reconstructive knee surgery.  

The data is in the file anesthesia.rda.

### Part 1a

Use R to create a two-way table with the type of anesthetic defining the rows and nausea after the surgery as the columns and also produce the output for a chi-square test for independence.

Is there an association between these two categorical variables at a 5% level of significance?

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
require(DS705data, quietly = T)
data("anesthesia")

table1 <- table(anesthesia$anesthetic, anesthesia$nausea)
table1
chisq.test(table1)

```

$H_0$: Anesthetic type and post surgery nausea are independent.

$H_a$: Anesthetic type and post surgery nausea are associated.

Reject $H_0$ at $\alpha = 0.05$, P=0.004535

There is sufficient evidence in this sample to conclude that the anesthetic type is associated with post surgery nausea (P=0.004535).

### Part 1b

Obtain the output from R (including the Wald tests for coefficients - so use "summary" function) for the logistic regression model with nausea as the dependent variable and the type of anesthetic as the predictor variable.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1b -|-|-|-|-|-|-|-|-|-|-|-

```{r}

mod_1 <- glm(nausea ~ anesthetic, data=anesthesia, family='binomial')
summary(mod_1)

anesthesia$method2 <- ifelse(anesthesia$nausea=='Yes',1,0)
summary(glm(method2 ~ anesthetic, data=anesthesia, family='binomial'))

# confirming that glm() converts 'Yes' to 1, and 'No' converted to 0 -> alphabetic order
# for anesthesia type, A will be 0 and B will be 1
```


### Part 1c  

What is the outcome of the hypothesis test that the coefficient of **anesthetic** is "zero" vs "not zero" at a 5% level of significance? (use the Wald test from the R output from the logistic regression you performed)

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1c -|-|-|-|-|-|-|-|-|-|-|-

Reject $H_0$ at $\alpha = 0.05$, P=0.00271. 

There is evidence to conclude that the coefficient of **anesthetic** is different from zero.

### Part 1d

Convert the estimated coefficient of **anesthetic** to an odds ratio and interpret it in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1d -|-|-|-|-|-|-|-|-|-|-|-

```{r}
exp(mod_1$coefficients[2])

```

The odds of having nausea under anesthetic B is `r round(exp(mod_1$coefficients[2])*100,2)`% as large as the odds of having nausea under anethetic A. That is, the odds under anesthetic B are `r 100-round(exp(mod_1$coefficients[2])*100,2)`% less.

### Part 1e

Install the package "mosaic" (if you don't have it installed already), then load it.  Use the oddsRatio function to compute the odds ratio for having nausea for anesthetic A vs B.  You may have to refer back to Week 8 for details on odds ratios and the oddsRatio function in R.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e -|-|-|-|-|-|-|-|-|-|-|-

```{r}
require(mosaic, quietly = T)
# A vs B, so we want A in the 2nd row (numerator in calculation)
# we want relative to a 'Yes' result, so 'Yes' is in column 1
oddsRatio(table1[c('B','A'),c('Yes','No')], verbose = T)
```


### Part 1f

When logistic regression coefficients are negative, the interpretation sometimes has more impact when we switch the perspective and use the reciprocal of the exponentiated coefficient.  Find the odds ratio for having nausea for anesthetic B compared to anesthetic A (i.e. take the reciprocal of the odds ratio you computed in part **1d**). 

Interpret this odds ratio in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1f -|-|-|-|-|-|-|-|-|-|-|-

```{r}
oddsRatio(table1[c('A','B'),c('Yes','No')]) # this is definitely B compared to A

odds1f <- exp(mod_1$coefficients[2]) # this is from 1d, looks like B compared to A to me
odds1f
```

The odds of having nausea under anesthetic B is `r round(odds1f*100,2)`% as large as the odds of having nausea under anethetic A. That is, anesthetic B is `r 100-round(odds1f*100,2)`% less.

Considering the reciprocal of the exponentiated coefficient, we see that odds of having nausea under anesthetic A is `r round(1/odds1f,2)` times larger than under anesthetic B.

### Part 1g
    
Compute the predicted probability of a reconstructive knee surgery patient having nausea after surgery when anesthetic A was used.
    
### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g -|-|-|-|-|-|-|-|-|-|-|-

```{r}
newData <- data.frame(anesthetic='A')
predict(mod_1, newdata = newData, type='response')

```

----

### Part 1h
    
Compute a 95% confidence interval for the predicted probability of a reconstructive knee surgery patient having nausea after surgery when anesthetic A was used.
    
### -|-|-|-|-|-|-|-|-|-|-|- Answer 1h -|-|-|-|-|-|-|-|-|-|-|-

```{r}
out <- predict(mod_1, newdata = newData, se.fit = T)
C <- 0.95
crit <- qnorm(1-(1-C)/2)
lower <- exp(out$fit-crit*out$se.fit)/(1+exp(out$fit-crit*out$se.fit))
upper <- exp(out$fit+crit*out$se.fit)/(1+exp(out$fit+crit*out$se.fit))
c(lower, upper)
```

## Exercise 2

Continue using the anesthesia.rda data set to do the following.

### Part 2a

Obtain the output from R (including the Wald tests for coefficients - so use "summary" function) for the logistic regression model with nausea as the dependent variable and the amount of pain medication taken as the predictor variable.  

At $\alpha = 0.05$, is there a statistically significant relationship between nausea and the amount of pain medication taken?

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
mod_2 <- glm(nausea ~ painmed, data=anesthesia, family='binomial')
summary(mod_2)
```

There is a statistically significant relationship between nausea and the amount of pain medication taken, at $\alpha = 0.05$, P=2.20e-05.

### Part 2b  

Convert the estimated coefficient of **painmed** to an odds ratio and interpret it in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
exp(mod_2$coefficients[2])
```


The odds of having nausea after reconstructive knee surgery increases by `r (1-round(odds1f*100,2))*100`% for every unit increase of pain medication.

### Part 2c

Compute the predicted probabilities of a reconstructive knee surgery patient having nausea in the recovery time after surgery for when 50 units of pain medication are used and also for when 100 units of pain medication are used.

Comment on these two probabilities.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2c -|-|-|-|-|-|-|-|-|-|-|-

```{r}
newData <- data.frame(painmed=c(50,100))
results.2c <- predict(mod_2, newdata = newData, type='response')
results.2c
```

The predicted probabilities of a reconstructive knee surgery patient having nausea in the recovery time after surgery when `r newData$painmed[1]` and `r newData$painmed[2]` units of pain medication are used are `r round(results.2c[1],2)` and `r round(results.2c[2],2)`, respectively.
