---
title: 'Two-Sample Inference Procedures for Population Central Values'
author: "Adam Hendel"
date: "02/13/2018"
output: word_document
fontsize: 12pt
---

```{r include=FALSE, cache=FALSE}
# Don't modify this chunk of code, it is just installing and loading the DS705data package
if (!require(DS705data)){
  if (!require(devtools)){
    install.packages('devtools',repos="http://cran.rstudio.com")
  }
  library(devtools)
  install_github('DataScienceUWL/DS705data')
}
require(DS705data)
```
   
## Exercise 1  

An experiment was conducted to evaluate the effectiveness of a drug treatment for tapeworm in the stomachs of sheep. A random sample of 24 worm-infected lambs of approximately the same age was randomly divided into two groups. Twelve of the lambs were injected with the drug and the remaining twelve were left untreated. After a 6-month period the lambs were slaughtered and the worm counts recorded.

### Part 1a

Load the dataset WormSheep from the DS705data package. Note that the package is loaded above at line 18.  You just need the data() command.  To see all the data sets in the package, type data(package='DS705data').

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# load dataset
data("WormSheep")
# inspect the data
str(WormSheep)
```

We can see one column with labels and one column with values--thus the dataset is 'stacked'.

---

### Part 1b	

Create boxplots and histograms for each group (treated vs. untreated).  Be sure that each plot is labeled appropriately.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
library(ggplot2)
# assign base ggplot object
base <- ggplot(WormSheep, aes(fill = treated), colour ='black') +
  theme(plot.title = element_text(hjust = 0.5))
# display boxplot
base + 
  geom_boxplot(aes(x=treated, y=worms)) +
  xlab('') +
  ylab('Number of Worms') +
  ggtitle('Boxplot of Worms in Sheep')
# display histogram
base + 
  geom_histogram(aes(x=worms), colour='black',bins=5) +
  facet_wrap(~treated) +
  xlab('Number of Worms') +
  ggtitle('Histogram of Worms in Sheep')
```

---

### Part 1c  

Do the boxplots show any outliers? 

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1c -|-|-|-|-|-|-|-|-|-|-|-

Per visual inspection of the boxplots, there are no outliers.

---

### Part 1d  

Describe the shapes of the histograms for the sample data for each sample.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1d -|-|-|-|-|-|-|-|-|-|-|-

Treated: The distribution is approximately normal, though it is a bit flatter that I'd like to see.

Untreated: There is a slight left-skewness to the distribution, though I would still say it has the general shape of a normal distribution. 

---

### Part 1e  

Do either of the normal probability plots show evidence of non-normality?  Explain.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1d -|-|-|-|-|-|-|-|-|-|-|-

The treated population is likely normal despite its somewhat flat appearance. The untreated population is slioghtly left-skewed, though it is not severe. I would not conclude non-normality for either population based on these histograms alone.

---

### Part 1f 

Conduct an appropriate test to determine if the worm counts in each population can be considered as normally distributed.  Provide the p-value and the conclusion of the test at a 5% level of significance.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1f -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# use Shapiro-Wilk to test if the samples were drawn from a normally distributed population.
# test the treated population
shapiro.test(WormSheep$worms[WormSheep$treated=='treated'])

```

Do not reject the null hypothesis at $a$=0.05, $P$=0.6504.

There is sufficient evidence to show that the treated sample was drawn from a normally distributed population.

```{r}
# test the untreated population
shapiro.test(WormSheep$worms[WormSheep$treated=='untreated'])
```

Do not reject the null hypothesis at $a$=0.05, $P$=0.5491.

There is sufficient evidence to show that the untreated sample was drawn from a normally distributed population.

---

### Part 1g	
    
Conduct the test of your choice to determine if the population mean worm count for all sheep treated with the drug is less than the mean worm count for the population of untreated sheep.  Let $\alpha = 0.05$.

#### Step 1

State the null and alternative hypotheses for the test.  If you use symbols, please define them.
 
#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g.step1 -|-|-|-|-|-|-|-|-|-|-|-

$$H_0: \mu_\mbox{treated} = \mu_\mbox{untreated}$$
$$H_a: \mu_\mbox{treated} < \mu_\mbox{untreated}$$


#### Step 2

Use R to generate the output for the test you selected.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g.step2 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# t.test on `stacked` data
t.test(worms~treated, data=WormSheep, alternative='less')
```

#### Step 3

State both a statistical conclusion at $\alpha=0.05$ and interpret it in the context of the problem.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g.step3 -|-|-|-|-|-|-|-|-|-|-|-

Reject the null hypothesis at $\alpha = 0.05$, P=0.02.

There is sufficient evidence to conclude that the population mean untreated worm count is greater than the population mean treated worm count.

---

### Part 1h  

Compute a 95% confidence interval for the difference in population means.  Write an interpretation in the context of the problem for this interval.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1h -|-|-|-|-|-|-|-|-|-|-|-

```{r}
t.test(worms~treated, data=WormSheep, conf.int=0.95)$conf
```

With 95% confidence, the population mean worm count for all treated sheep is 1.13 to 25.03 worms less than for all untreated sheep.

---

### Part 1i  

Did you use the Welch t-test or the Wilcoxon rank sum test?  Justify your choice, including some discussion of how well the conditions for the hypothesis test and confidence interval procedures were met.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1i -|-|-|-|-|-|-|-|-|-|-|-

I chose the Welch t-test due to the distributions of the sample datasets being close to approximately normal. Neither the treated nor untreated samples had heavy skewness by visual inspection of the histograms. Neither samples had large outliers by inspection of the boxplots. The numerical test for normality, Shapiro-Wilk, also indicated that these samples could have been drawn from a normal distribution. These conditions set the stage nicely for usage of the Welch t-test--if there were large outliers and heavy skewness, the Wilcoxon Rank Sum approach would have been more appropriate.

---

## Problem 2

Consider two independent, random samples of delay times (in hour) for two airlines over the past 10 years. The task here is to compare the population central values by conducting a hypothesis test and constructing a 95% confidence interval for the difference in average delay time between the airlines for the population of all delays in the past 10 years.

### Part 2a  

Load the dataset delays from the DS705data package.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# load data and inspect
data("delays")
str(delays)
```

---

### Part 2b

Begin by exploring the sample means, medians, and standard deviations for each airline. That is, find the sample means, medians, and standard deviations. 

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
airlines <- unique(delays$airline)
for(airline in airlines){
  means <- mean(delays$delay[delays$airline==airline])
  medians <- median(delays$delay[delays$airline==airline])
  s.d. <- sd(delays$delay[delays$airline==airline])
  print(paste(airline, 'mean:', round(means,2), 'median:', round(medians,2), 'sd:', round(s.d.,2), sep=' '))
}
```

---

### Part 2b

Produce boxplots, histograms, and normal probability plots for the delay times of both airlines.  

Discuss the aspects of the plots that are relevant to the construction of the best type of test and confidence interval to compare the population central values.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# base ggplot object
base <- ggplot(delays)
# display boxplots
base +
  geom_boxplot(aes(x=airline, y=delay))
# display histograms
base +
  geom_histogram(aes(x=delay), bins=9, fill='gray', col='black') +
  facet_wrap(~airline)
# display normal probability plots
qqnorm(delays$delay[delays$airline=='Omega'], main='Normal QQ - Omega')
qqline(delays$delay[delays$airline=='Omega'])

qqnorm(delays$delay[delays$airline=='Skybird'], main='Normal QQ - Skybird')
qqline(delays$delay[delays$airline=='Skybird'])
```

We see both airlines have outliers on the upper end of their distributions by visually inspecting the boxplots. The histograms and qqplots indicate right skewness as well. Due to the outliers and skewness of the distributions, either the Bootstrap or Wilcoxon Rank Sum method would be more appropriate than the t methods.

---

### Part 2c

Conduct hypothesis tests for the normality of the delay times for both Skybird and Omega using a 5% level of significance in each case.

Use R to compute the test statistic and P-value.  

State the hypotheses and conclusion of the test. Include the P-value in your conclusion.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2c -|-|-|-|-|-|-|-|-|-|-|-

H~0~ The sample was drawn from a normally distributed population.

H~a~ The sample was not drawn from a normally distributed population.

```{r}
# test Omega
shapiro.test(delays$delay[delays$airline=='Omega'])
# test Skybird
shapiro.test(delays$delay[delays$airline=='Skybird'])
```

Omega: Reject the null hypothesis at $\alpha = 0.05$, P=6.027e-05.

There is sufficient evidence to conclude that the Omega sample was not drawn from a normally distributed population

Skybird: Reject the null hypothesis at $\alpha = 0.05$, P=0.0006031.

There is sufficient evidence to conclude that the Skybird sample was not drawn from a normally distributed population

---

### Part 2d	

Perform the most appropriate test to compare population central values for all Skybird and Omega delay times for the past 10 years.  Use a 10% level of significance.

#### Step 1

State the null and alternative hypotheses for the test (in words).
 
#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2d.step1 -|-|-|-|-|-|-|-|-|-|-|-

H~0~ The population mean of Omega flight delays is equal to Skybird's.

H~a~ The population mean of Omega flight delays is greater than Skybird's.

#### Step 2

Use R to generate the output for the test you selected.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2d.step2 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# condut the test
wilcox.test(delay~airline, data=delays)
```

#### Step 3

State both a statistical conclusion at $\alpha=0.1$ and interpret it in the context of the problem.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2d.step3 -|-|-|-|-|-|-|-|-|-|-|-

Reject the null hypothesis at $\alpha=0.1$, P=0.8173.

There is not sufficient evidence to conclude that the population mean flight delay for Omega is greater than Skybird's.

---

### Part 2e

Construct the most appropriate 95% confidence interval to compare population central values.  Write a sentence to interpret the interval in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2e -|-|-|-|-|-|-|-|-|-|-|-

```{r}
wilcox.test(delay~airline, data=delays, conf.int=0.95)$conf.int
```

With 95% confidence, the population mean flight delay of Omega is the same as Skybird's.

---

### Part 2f

Did you use the Welch t-test and interval or the Wilcoxon rank sum test with the corresponding interval?  Justify your choice, including some discussion of how well the conditions for the hypothesis test and confidence interval procedures were met.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2f -|-|-|-|-|-|-|-|-|-|-|-

I selected the Wilcoxon Rank Sum and its confidence interval. This decision was based on the evidence from the diagnostic plots that were analyzed prior to the test. The presence of strong outliers and skewness and general non-normality of the sample distributions lead me choosing the Wilcoxon method over the Welch t-test.

---

### Part 2g

What type of hypothesis testing error could have been made here?  Describe it in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2g -|-|-|-|-|-|-|-|-|-|-|-

Since we chose not to reject the null hypothesis, the chance for a type II error exists. We would be in a type II error situation if the true population means were actually different, though we did not reject the hypothesis that they are the same. We'd have a type two error if, for example, the population mean delay times of Omega turned out to be statistically different than Skybird's.

---

### Part 2h

Construct a 95% bootstrap confidence interval to compare population medians for the delay times over the past 10 years for the two airlines. Use the Bca method and a bootstrap sample size of 2000. Compare it to the interval produced by the wilcox.test function.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2g -|-|-|-|-|-|-|-|-|-|-|-

```{r}
library(boot)
bootMeanDiff <- function(d,i){
  means <- tapply(d[i,1],d[,2],mean)
  means[1]-means[2]
}

boot.object <-boot(delays, bootMeanDiff, R = 2000, strata=delays$airline)
boot.ci(boot.object, type='bca',conf = 0.95)$bca[4:5]
```

With 95% confidence, the population mean delay times for Omega are equal to Skybird's.

The CI results from the bootstrap are similar to wilcox.test function. Wilcox resulted in difference of -.39 to 0.28 while the bootstrap method resulted in -.41 to .38 difference. 